---
title: "Homework 1"
author: "Yi Zuo"
date: "`r date()`"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('ElemStatLearn')

## load prostate data
data("prostate")

## subset to training examples
prostate_train <- subset(prostate, train=TRUE)

```

Using the RMarkdown/knitr/github mechanism, implement the following tasks by extending the example R script (prostate-data-lin.R):  

* Write functions that implement the L1 loss and tilted absolute loss functions.    

```{r}

## L2 loss function
L2_loss <- function(y, yhat) (y-yhat)^2

## L1 loss function
L1_loss <- function(y, yhat) abs(y-yhat)

# tilted absoulte loss function
t_loss <- function(y, yhat, tau) {
  tau * abs(y-yhat) * (+(y>yhat)) + (tau - 1) * (y-yhat) * (+(y<=yhat))}

```

* Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the linear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.    

```{r}

###############
#   L2 loss   #
###############

## fit simple linear model using numerical optimization
fit_lin_l2 <- function(y, x, beta_init = c(-0.51, 0.75)) {
  err <- function(beta)
    mean(L2_loss(y,  beta[1] + beta[2]*x))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

###############
#   L1 loss   #
###############

## fit simple linear model using numerical optimization
fit_lin_l1 <- function(y, x, beta_init = c(-0.51, 0.75)) {
  err <- function(beta)
    mean(L1_loss(y,  beta[1] + beta[2]*x))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

############################
#   tilted absolute loss   #
############################

## fit simple linear model using numerical optimization
fit_lin_t <- function(y, x, tau, beta_init = c(-0.51, 0.75)) {
  err <- function(beta)
    mean(t_loss(y,  beta[1] + beta[2]*x, tau))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

## make predictions from linear model
predict_lin <- function(x, beta) beta[1] + beta[2]*x

## fit linear model
lin_beta_l2 <- fit_lin_l2(y=prostate_train$lcavol,
                    x=prostate_train$lpsa)

lin_beta_l1 <- fit_lin_l1(y=prostate_train$lcavol,
                    x=prostate_train$lpsa)

lin_beta_t_0.25 <- fit_lin_t(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,tau=0.25)

lin_beta_t_0.75 <- fit_lin_t(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,tau=0.75)

## compute predictions for a grid of inputs
x_grid <- seq(min(prostate_train$lpsa),
              max(prostate_train$lpsa),
              length.out=100)

###############
#   L2 loss   #
###############

lin_pred_l2 <- predict_lin(x=x_grid, beta=lin_beta_l2$par)

###############
#   L1 loss   #
###############

lin_pred_l1 <- predict_lin(x=x_grid, beta=lin_beta_l1$par)

############################
#   tilted absolute loss   #
############################

lin_pred_t_0.25 <- predict_lin(x=x_grid, beta=lin_beta_t_0.25$par)

lin_pred_t_0.75 <- predict_lin(x=x_grid, beta=lin_beta_t_0.75$par)

## plot data
plot(prostate_train$lpsa, prostate_train$lcavol,
       xlab="log Prostate Screening Antigen (psa)",
       ylab="log Cancer Volume (lcavol)",main="Linear models")

## plot predictions
lines(x=x_grid, y=lin_pred_l2,col="red")
lines(x=x_grid, y=lin_pred_l1,col="brown")
lines(x=x_grid, y=lin_pred_t_0.25,col="blue")
lines(x=x_grid, y=lin_pred_t_0.75,col="green")
legend("bottomright",legend=c("L2 loss","L1 loss","Tilted 0.25", "Tilted 0.75"),col=c("red","brown","blue","green"),cex=0.8,lty=1)

```


* Write functions to fit and predict from a simple exponential (nonlinear) model with three parameters defined by 'beta[1] + beta[2]*exp(-beta[3]*x)'. Hint: make copies of 'fit_lin' and 'predict_lin' and modify them to fit the nonlinear model. Use c(-1.0, 0.0, -0.3) as 'beta_init'.  

```{r}

###############
#   L2 loss   #
###############

## fit simple linear model using numerical optimization
fit_nlin_l2 <- function(y, x, beta_init = c(-1.0, 0.0, -0.3)) {
  err <- function(beta)
    mean(L2_loss(y,  beta[1] + beta[2]*exp(-beta[3]*x)))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

###############
#   L1 loss   #
###############

## fit simple linear model using numerical optimization
fit_nlin_l1 <- function(y, x, beta_init = c(-1.0, 0.0, -0.3)) {
  err <- function(beta)
    mean(L1_loss(y,  beta[1] + beta[2]*exp(-beta[3]*x)))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

############################
#   tilted absolute loss   #
############################

## fit simple linear model using numerical optimization
fit_nlin_t <- function(y, x, tau, beta_init = c(-1.0, 0.0, -0.3)) {
  err <- function(beta)
    mean(t_loss(y,  beta[1] + beta[2]*exp(-beta[3]*x), tau ))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

```


* Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the nonlinear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.   

```{r}

## make predictions from linear model
predict_nlin <- function(x, beta) beta[1] + beta[2]*exp(-beta[3]*x)

## fit linear model
nlin_beta_l2 <- fit_nlin_l2(y=prostate_train$lcavol,
                    x=prostate_train$lpsa)

nlin_beta_l1 <- fit_nlin_l1(y=prostate_train$lcavol,
                    x=prostate_train$lpsa)

nlin_beta_t_0.25 <- fit_nlin_t(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,tau=0.25)

nlin_beta_t_0.75 <- fit_nlin_t(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,tau=0.75)

## compute predictions for a grid of inputs
x_grid <- seq(min(prostate_train$lpsa),
              max(prostate_train$lpsa),
              length.out=100)

###############
#   L2 loss   #
###############

nlin_pred_l2 <- predict_nlin(x=x_grid, beta=nlin_beta_l2$par)

###############
#   L1 loss   #
###############

nlin_pred_l1 <- predict_nlin(x=x_grid, beta=nlin_beta_l1$par)

############################
#   tilted absolute loss   #
############################

nlin_pred_t_0.25 <- predict_nlin(x=x_grid, beta=nlin_beta_t_0.25$par)

nlin_pred_t_0.75 <- predict_nlin(x=x_grid, beta=nlin_beta_t_0.75$par)

## plot data
plot(prostate_train$lpsa, prostate_train$lcavol,
       xlab="log Prostate Screening Antigen (psa)",
       ylab="log Cancer Volume (lcavol)",main="Nonlinear models")

## plot predictions
lines(x=x_grid, y=nlin_pred_l2,col="red")
lines(x=x_grid, y=nlin_pred_l1,col="brown")
lines(x=x_grid, y=nlin_pred_t_0.25,col="blue")
lines(x=x_grid, y=nlin_pred_t_0.75,col="green")
legend("bottomright",legend=c("L2 loss","L1 loss","Tilted 0.25", "Tilted 0.75"),col=c("red","brown","blue","green"),cex=0.8,lty=1)

```

