---
title: "Homework 2"
author: "Yi Zuo"
date: "`r date()`"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Using the RMarkdown/knitr/github mechanism, implement the following tasks by extending the example R script (prostate-data-lin.R):  

* Use the prostate cancer data from the `ElemStatLearn` package for R.

```{r}
library('ElemStatLearn')

## load prostate data
data("prostate")

## subset to training examples
prostate_train <- subset(prostate,train==T)
prostate_test <- subset(prostate,train==F)

```

* Use the `cor` function to reproduce the correlations listed in HTF Table 3.1 page 50

```{r}
round(cor(prostate_train[,1:(ncol(prostate)-1)]),3)
```

* Treat `lpsa` as the outcome, and use all other variables in the data set as predictors  

* With the training subset, train a least-squares regression model with all predictors using the `lm` function (with the training subset)  

```{r}
ls <- lm(lpsa~.,data=prostate_train[,-ncol(prostate_train)])

```

* Use the testing subset to compute the test error using the fitted least-squares regression model  

The test error for each observation in the test set is stored in `te`.  

```{r}
te <- prostate_test$lpsa - predict(ls,newdata=prostate_test[,c(-9,-10)])
te
```

# Train a ridge regression model using the `glmnet` function, and tune the value of `lambda`  

```{r}
library(glmnet)



```



